Stochastic blockmodeling of relational event dynamics

### 2012-06-01 docs/intro.tex
Small paper edits.  A few TODOs inserted.

In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.  While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.  Consider, e.g., Figure~\ref{fig:introexample}.  In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.  As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.  By contrast, consider the interaction patterns shown in the top right panel.  Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.

>>>>

In the cross-sectional case, \emph{stochastic blockmodels} \cite{Nowicki2001, Kemp2006, Ishiguro2010, Rodriguez2011} have been proposed as a family of approaches that capture behavioral similarity by identifying subsets of nodes with similar patterns of ties to those in other sets.  While it is natural to apply these ideas directly to relational event data via blockmodeling of the time-marginalized rates of interaction among dyads (effectively treating the event structure as a valued graph), there are limits to what this approach can detect.  Consider, e.g., Figure~\ref{fig:introexample}.  In the top left panel, we depict the time-marginalized frequencies of simulated interactions between members of two groups (A and B), with darker cells indicating higher interaction frequencies.  As can be seen, a clear role structure is present, with members of each subgroup interacting at higher rates with co-members than out-group members; such a structure is easily detectable via conventional blockmodeling techniques.  By contrast, consider the interaction patterns shown in the top right panel.  Here, there is no systematic difference in marginal rates between the two groups, rendering them invisible under a standard blockmodeling approach.

### 2012-06-01 docs/model.tex
Small paper edits.  A few TODOs inserted.

We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
%z_i &\sim \mbox{CRP}(\alpha) &
%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
Thus, our model for the rate of $(i,j)$ events has the same parameters as for other events occurring between group $z_i$ and $z_j$.

 In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.
% TODO: Lead off with computational convenience, then justify?
%\footnote{TODO: This is analogous to a Markov condition for this class of models.}

We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta{\sigma})$.
The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
% TODO: Expand on this?

@@

In Table \ref{tab:stats} we list the statistics we use in our experiments for  $\mathbf{s}(t,i,j)$.
For example, a particular node sending often may indicate they will continue to be a sender in the future.
We normalize these counts by the number of events up until a dyad's prior changepoint. %\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.}
Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

% TODO: Awkward next sentence
The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
Though we use only the statistics in Table \ref{tab:stats} in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
% TODO: Also, need to be restricted to ego conditions if we want it fast

>>>>

{\color{red} (Need a better transition here.)}  We aim to learn about the dynamics between subsets of nodes.   To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
%z_i &\sim \mbox{CRP}(\alpha) &
%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
Thus, the rate of $(i,j)$ events has the same parameters as other events occurring between group $z_i$ and $z_j$.

% In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.

We allow the blocks to share information by placing a hierarchical prior on the collection of $\boldsymbol{\beta}_{k.l}$ where $\mu_p \propto 1$ and $\sigma_p^2 \sim \mbox{Inv-Gamma}(\alpha_{\sigma},\beta_{\sigma})$.
The cluster assignments are given a non-parameteric prior $z_i \sim \mbox{CRP}(\alpha)$.
% TODO: Expand on this?

@@

Table \ref{tab:stats}  lists the statistics  $\mathbf{s}(t,i,j,\mathcal{A}_t)$ used in Section \ref{sec:experiments}.
For example, a particular node sending often may indicate they will continue to be a sender in the future.
We normalize these counts by the number of events up until a dyad's prior changepoint. %\footnote{This seems to avoid ``explosion'' when simulating from the model.  TODO: Explore the theory about these sorts of statistics.  Aalen has a few suggestions.}
Other statistics could be of interest for particular substantive questions \cite{Butts2008,Vu2011}.  \footnote{Describe the form of $f()$.  Right now it is $f(x) = \log \frac{x+1}{m + N(N-1)}$.}

% TODO: Awkward next sentence
The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
The only restriction is that the statistic may not change in value between each observed event.\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
% TODO: Also, need to be restricted to ego conditions if we want it fast

### 2012-06-01 docs/simulation.tex
Small notation fixes.
Small paper edits.  A few TODOs inserted.

\begin{figure}
%\centering
\begin{subfigure}[b]{0.25\textwidth}
\centering
\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf}
\end{subfigure}
~
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
\end{subfigure}
~
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
\end{subfigure}
\caption{Illustration of 2000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.  Right: Parameters (in red) and posterior credible intervals (in black).}
\label{fig:syncounts}
\end{figure}

@@

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 clusters where 1) the first cluster has an increased tendency for reciprocity, 2) members of the second cluster have an increased tendency to subsequently talk to someone within their cluster.
The specification of $\textbf{s}$ is therefore $s(t,i,j) = [s_0, s_{1}(t,i,j), s_{2}(t,i,j)]$.
For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
This is done by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.

>>>>

\begin{figure}
%\centering
\begin{subfigure}[b]{0.25\textwidth}
\centering
\includegraphics[width=1.5in]{../figs/synthetic/mat.pdf}
\end{subfigure}
~
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=2in]{../figs/synthetic/counts.pdf}
\end{subfigure}
~
\begin{subfigure}[b]{0.3\textwidth}
\centering
\includegraphics[width=2.5in]{../figs/synthetic/params-estimates.pdf}
\end{subfigure}
\caption{[TODO: Remove AB-XA, etc from the plot. Rotate them. Use subfigure labels.] Illustration of 2000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2.  Right: Parameters (in red) and posterior credible intervals (in black).}
\label{fig:syncounts}
\end{figure}

@@

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 groups where 1) within group communication is more likely, 2) events among members in the first group are more likely to be reciprocated  (i.e. a positive \texttt{ab-ba} effect), and 3) events among members of the second group are more likely to be followed by an event with the same sender (i.e. a positive \texttt{ab-ay} effect).  The specification of $\textbf{s}$ is therefore $\textbf{s}(t,i,j,\mathcal{A}_t) = [s_0, s_{1}(t,i,j,\mathcal{A}_t), s_{2}(t,i,j,\mathcal{A}_t)]$.  For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
Data is generated by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.
